\subsection{Practical: Linear Advection}

Write your answers on this piece of paper.

In the subdirectory \url{pythonExamples/advection} you will find code for solving the linear advection equation:
\begin{equation*}
\frac{\partial\phi}{\partial t} + u\frac{\partial\phi}{\partial x} = 0
\end{equation*}
with cyclic boundaries conditions. The code is organised as:\\
\begin{tabular}{ll}
\url{linearAdvect.py} & outer code to define input variables, call linear\\
                      & advection functions and plot the results \\
\url{initialConditions.py} & code to generate the initial conditions \\
\url{advectionSchemes.py} & Code defining CTCS and FTBS schemes. \\
                          & This is where you should add code for other schemes.
\end{tabular}

The following setup variables are defined in the outer code:\\
\begin{tabular}{lcl}
Start location & $x_{\min}$ & 0\\
End location & $x_{\max}$ & 1m\\
Number of points & $n_x$ & 100\\
Spatial step & $\Delta x$ & $(x_{\max} - x_{\min})/n_x$ \\
Courant number & $c$ & 0.2 \\
Wind speed & $u$ & 1m/s\\
Time step & $\Delta t$ & $c ~ \Delta x/u$ \\
Number of time steps & $n_t$ & 50\\
\end{tabular}

\begin{enumerate}
\item Run the code and identify the diffusion errors and dispersion errors of the two schemes.
\\ \optparagraph{
FTBS suffers from diffusion errors\\
CTCS suffers from dispersion errors
}

%\clearpage
\item Some error norms of numerical results are defined as:\\
\begin{tabular}{ccc}
$\ell_1(\phi(x)) = \frac{\int_x |\phi - \phi_t| dx}{\int_x |\phi_t| dx}$
&
$\ell_2(\phi(x)) = \sqrt{\frac{\int_x (\phi - \phi_t)^2dx}{\int_x \phi_t^2dx}}$
&
$\ell_\infty(\phi(x)) = \frac{\max|\phi - \phi_t|}{\max|\phi_t|}$
\end{tabular}\\
for a numerical solution, $\phi(x)$, in comparison to the exact solution, $\phi_t$. The code in \url{linearAdvect.py} calculates these error norms as well as the mean and standard deviation of the initial and final solutions. For different resolutions (change $n_x$ and change $n_t$ so as to get the same end time) fill in the table for the error norms from each scheme.

\hspace{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|}
\hline
$n_x$ & \multicolumn{2}{c|}{Initial} & \multicolumn{4}{c|}{FTBS} & \multicolumn{4}{c|}{CTCS} \\
 & mean & std. dev & mean & std. dev & $\ell_2$ & $\ell_\infty$ & mean & std. dev & $\ell_2$ & $\ell_\infty$ \\
\hline
50 & \opttext{0.384} & \opttext{0.401} & \opttext{0.384} & \opttext{0.341} & \opttext{0.023} & \opttext{0.283} & \opttext{0.384} & \opttext{0.401} & \opttext{0.011} & \opttext{0.219}\\
\hline
100 & \opttext{0.379} & \opttext{0.400} & \opttext{0.379} & \opttext{0.367} & \opttext{0.009} & \opttext{0.219} & \opttext{0.379} & \opttext{0.400} & \opttext{0.002} & \opttext{0.087}\\
\hline
200 & \opttext{0.378} & \opttext{0.399} & \opttext{0.377} & \opttext{0.382} & \opttext{\footnotesize 0.0033} & \opttext{0.159} & \opttext{0.378} & \opttext{0.400} & \opttext{\footnotesize 0.00056} & \opttext{0.071} \\
\hline
\end{tabular}

\begin{enumerate}
\item Which of the schemes conserves the total quantity of mass?
\item Which of the schemes conserves the variance of the tracer?
\item Calculate the order of accuracy of the schemes based on the $\ell_2$ and $\ell_\infty$ norms. Are they as expected?
\end{enumerate}

\item Change the Courant number to 1.1 and run for 10 time-steps. What is happening to the solution? What is happening if you run for longer? Does the same problem occur for a Courant number of 1? What about a Courant number of -0.2? (What would this imply?)  Does the same problem occur if you remove the sharp profile from the initial conditions?
\\ \optparagraph{
The solution is diverging. It is not stable. \\
If you run for longer the solution will become infinite\\
For a Courant number less than or equal to 1 the solution is stable. \\
For a Courant number of less than zero, FTBS is not stable. This implies wind in the opposite direction\\
The same problems occur for just the smooth profile, but the instabilities do not occur so quickly.
}
\item Write down and implement one of the linear advection schemes CTBS (centred in time, backward in space) and FTCS (forward in time, centred in space). Can you find a Courant number for which these schemes are stable?
\\ \optparagraph{FTCS and CTBS are unstable for all Courant numbers}
\end{enumerate}

